Application of Generalized Additive Models for long-term and seasonal trend analysis of water quality: South San Francisco Bay case study

### 2020-09-23 manu_draft.Rmd
editing intro for completeness

Many environmental monitoring programs generate fine-scale but potentially irregular time-series data for the purpose of assessing long-term trends for regulatory, management, or scientific purposes.  The mismatch between the scales of monitoring versus analysis questions can lead to statistical challenges.  At finer temporal scales, environmental systems may show short-term fluctuations from multiple factors (e.g., weather events or seasonal changes) that are not of direct interest for longer-term trends.  For basic questions, simple smoothing or averaging over time may be appropriate for analyzing longer-term trends. However, simple analyses that aggregate or generalize observations to assess trends can create mismatches between questions of interest and what is returned by the summarized results.  Further, many analysis questions are not easily addressed by simple smoothing methods, such as an assessment of trends that occur at different temporal scales.

@@

GAMs have only recently been applied to trend analysis of water quality time-series, particularly in the evaluation of multi-decadal time series from long-term monitoring programs.  The Chesapeake Bay Program has developed and integrated GAM methods into routine reporting assessments to track and understand long-term changes in nutrients and other eutrophication indicators.  Central to their method is flexibility in describing the individual components of the time series with varying levels of complexity.  Post-hoc methods were also integrated into the GAM framework to address regionally-specific programmatic or management questions, such as an assessment of GAM fit before/after a management or policy intervention of interest (e.g., impoundment removal or point-source control).  Other studies have addressed specific technical issues of applying GAMs to environmental time series, such as appropriate transformations for response variables, accounting for back-transformation bias, addressing serial correlation in high resolution data, or addressing different time lags in describing relationships between response and predictor variables.

Although our approach is not the first to apply GAMs to environmental monitoring data, our approach is distinct from previous applications in several ways.  First, some approaches attempt to estimate a parameter such as a seasonal average with the GAM itself [q1].  This represents a mismatch between time scales and does not make sense as an estimation method if there is really no seasonal treatment factor, for example.  Second, some approaches formulate GAMs with multiple spline terms to decompose time-scales of fluctuations within the GAM fitting machinery [q2].  There may be some analysis questions for which this is appropriate, but it is less general than our approach and it goes against caution of Wood (2nd edition book) that the degrees of smoothness attributed to multiple splines can be hard to interpret.  Perhaps more important, it risks obscuring the different time-scales implicit in choices such as number of spline knots, leaving the rendering results harder to interpret.  Third…[q3]...  We illustrate each of these points in more detail below.

We illustrate the proposed methods with the motivating example of water quality monitoring in the southern portion of San Francisco Bay, California, USA.  For several decades, approximately twice-monthly monitoring has been conducted at fixed locations (stations) of the longitudinal axis of the Bay.  Measurements from water samples include chlorophyll a, [others].  Analysis is complicated by irregularities in timing and consistency of data collection, such that simple seasonal averages of raw data could be misleading.  Examples of long-term trend questions include: Are there significant trends in spring mean chl-a at four-year (or other) time-scales?  At what across-year scale do within-year summer-fall mean chl-a levels change?  Are there significant across-year trends in within-year timing of the spring bloom in chl-a or in off-bloom baseline levels?  We also contrast different GAM formulations that can yield similar signal-extraction results but carry somewhat different interpretations.  The approach we give for using meta-analysis methods as a second stage of analysis following signal extraction with GAMs appears to be new to environmental trend-detection problems.  For this step, we give methods for isolating seasonal trends secondarily from GAM results with reasonable certainty and evlauating these trends between year.  Practically, this approach answers if a seasonal change has occurred across years of interest by providing an estimate of the change and its associated confidence as propagated through the GAM.

>>>>

Many environmental monitoring programs generate fine-scale but potentially irregular time-series data for the purpose of assessing long-term trends for regulatory, management, or scientific purposes.  The mismatch between the scales of monitoring versus analysis questions can lead to statistical challenges.  At finer temporal scales, environmental systems may show short-term fluctuations from multiple factors (e.g., weather events or seasonal changes) that may not be of direct interest for longer-term trends.  For basic questions, simple smoothing or averaging over time may be appropriate for analyzing longer-term trends. However, simple analyses that aggregate or generalize observations to assess trends can create mismatches between questions of interest and what is returned by the summarized results.  Further, many analysis questions are not easily addressed by simple smoothing methods, such as an assessment of trends that occur at different temporal scales.

@@

GAMs have only recently been applied to trend analysis of water quality time-series, particularly in the evaluation of multi-decadal time series from long-term monitoring programs.  The Chesapeake Bay Program has developed and integrated GAM methods into routine reporting assessments to track and understand long-term changes in nutrients and other eutrophication indicators.  Central to their method is flexibility in describing the individual components of the time series with varying levels of complexity, including modeling additive annual and seasonal changes and the interaction between the two.  Post-hoc methods were also integrated into the GAM framework to address regionally-specific programmatic questions, such as an assessment of GAM fit before/after a management or policy intervention of interest (e.g., impoundment removal or point-source control).  Other studies have addressed specific technical issues of applying GAMs to environmental time series, such as appropriate transformations for response variables, accounting for back-transformation bias, addressing serial correlation in high resolution data, or addressing different time lags in describing relationships between response and predictor variables.

Although our approach is not the first to apply GAMs for trend analysis environmental monitoring data, our approach is distinct from previous applications in several ways.  First, some approaches use the GAM smoothing functions to approximate components of a time series to estimate average conditions for a finite period of time.  This can represent a mismatch between processes occurring at different time scales and does not make sense as an estimation method if, for example, a seasonal treatment factor is not an explicit component of the time series.  Second, some approaches formulate GAMs with multiple spline terms to decompose time-scales of fluctuations within the boundaries of the GAM estimation methods.  There may be some analysis questions for which this is appropriate, but it is less general than our approach and it goes against caution of Wood (2nd edition book) that the degrees of smoothness attributed to multiple splines can be confounded.  Perhaps more important, this approach risks obscuring the different time-scales implicit in choices such as an appropriate number of spline knots, leaving the rendering results more difficult to interpret.  Third, over-confidence in conclusions derived from insufficient treatment of model uncertainty, including propogation to secondary analyses, can lead to misinformed decisions.  GAM applications to environmental monitoring data have not fully treated error propagation as a fundamental component of model application to decision-making, particularly when wrong decisions can lead to expensive consequences.

We illustrate the proposed methods with the motivating example of water quality monitoring in the southern portion of San Francisco Bay, California, USA.  For several decades, approximately twice-monthly monitoring has been conducted at fixed locations (stations) of the longitudinal axis of the Bay.  Analysis of these data is complicated by irregularities in timing and consistency of data collection, such that simple seasonal averages of raw data may not adequately describe trends.  Examples of long-term trend questions include: Are there significant trends in spring mean chlorophyll at four-year (or other) time-scales?  At what across-year scale do within-year summer-fall mean chloropyll levels change?  Are there significant across-year trends in within-year timing of the spring bloom in chlorophyll or in baseline levels during periods of low productivity?  We also contrast different GAM formulations that can yield similar signal-extraction results but carry somewhat different interpretations.  The approach we give for using meta-analysis methods as a second stage of analysis following signal extraction with GAMs appears to be new to environmental trend-detection problems.  For this step, we give methods for isolating seasonal trends secondarily from GAM results with reasonable certainty and evaluating these trends between year.
